---
title: "PalantiR engine validation"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
params:
  seed: 20260821
  n_sites: 3000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, warning = FALSE, message = FALSE)
suppressMessages(library(PalantiR))
use_genetic_code("Standard nuclear")
set_palantir_seed(params$seed)
set.seed(params$seed)

verdicts <- list()
verdict <- function(id, ok, detail) {
    verdicts[[id]] <<- ok
    cat(sprintf("**%s: %s** — %s\n", id, ifelse(ok, "PASS", "FAIL"), detail))
}
has_logo <- requireNamespace("ggseqlogo", quietly = TRUE)

aa_of <- unname(GeneticCode())
aa_levels <- sort(unique(aa_of))
syn_mask <- outer(aa_of, aa_of, "==")
diag(syn_mask) <- FALSE

# Sequence logo of one or more amino-acid frequency distributions
# (columns of `freqs`), with a barplot fallback if ggseqlogo is absent.
plot_freqs <- function(freqs, labels) {
    if (has_logo) {
        m <- as.matrix(freqs); rownames(m) <- aa_levels; colnames(m) <- labels
        print(ggseqlogo::ggseqlogo(m, method = "custom", seq_type = "aa") +
              ggplot2::scale_x_continuous(breaks = seq_along(labels), labels = labels) +
              ggplot2::ylab("frequency"))
    } else {
        barplot(t(as.matrix(freqs)), beside = TRUE, names.arg = aa_levels,
                legend.text = labels, ylab = "frequency")
    }
}

codon_to_aa <- function(codon_counts) as.numeric(tapply(codon_counts, aa_of, sum))
tv <- function(p, q) 0.5 * sum(abs(p - q))
syn_outflux <- function(model) rowSums(model$transition * syn_mask)

# Independently integrate a selected process until it expends a specified dS
# clock budget. The augmented state tracks the codon distribution, the dS
# functional, and synonymous opportunity exposure. This avoids relying on the
# simulator's internal time-change implementation when constructing the test
# expectation.
rk4_to_budget <- function(p0, Q, h, g, budget, dt = 0.001) {
    deriv <- function(y) {
        p <- y[seq_along(p0)]
        c(as.numeric(p %*% Q), sum(p * h), sum(p * g))
    }
    step <- function(y, d) {
        k1 <- deriv(y)
        k2 <- deriv(y + d * k1 / 2)
        k3 <- deriv(y + d * k2 / 2)
        k4 <- deriv(y + d * k3)
        y + d * (k1 + 2 * k2 + 2 * k3 + k4) / 6
    }
    n <- length(p0)
    y <- c(p0, dS_clock = 0, syn_exposure = 0)
    tau <- 0
    while (y[n + 1] < budget) {
        proposal <- step(y, dt)
        if (proposal[n + 1] <= budget) {
            y <- proposal
            tau <- tau + dt
        } else {
            lo <- 0; hi <- dt
            for (i in seq_len(40)) {
                mid <- (lo + hi) / 2
                if (step(y, mid)[n + 1] < budget) lo <- mid else hi <- mid
            }
            y <- step(y, (lo + hi) / 2)
            tau <- tau + (lo + hi) / 2
        }
    }
    list(tau = unname(tau), dS_clock = unname(y[n + 1]),
         syn_exposure = unname(y[n + 2]))
}
```

This notebook tests the core simulation engine against exact expectations.
Each test states its expectation, runs a fresh simulation, compares the
result statistically and visually, and prints a PASS or FAIL verdict against
a criterion stated in advance. To rerun everything with a different seed or
simulation size:

```r
rmarkdown::render("engine_validation.Rmd", params = list(seed = 1, n_sites = 3000))
```

The stationary-distribution and generic-transient tests use fitness profiles
shipped with the package (`psi_c50_1`). The dS tests additionally construct a
standard Zhang--Nielsen--Yang branch-site model and a mutation--selection
fitness and population-size shift.

```{r inputs}
data(psi_c50_1)
hky <- HasegawaKishinoYano(equilibrium = rep(0.25, 4))
psi_1 <- equilibrium_to_fitness(as.numeric(psi_c50_1[1, ]), 1000)
psi_2 <- equilibrium_to_fitness(as.numeric(psi_c50_1[2, ]), 1000)
n_sites <- params$n_sites
```

# Test 1: branch-length calibration under event-count scaling

**Expectation.** A model built with `scaling_type = "synonymous-per-codon"`,
`"non-synonymous"`, or `"substitution"` is normalized so that one unit of
branch length delivers, at stationarity, one expected substitution of that
class per codon site: synonymous, non-synonymous, or of any kind,
respectively. Simulating `r sprintf("%d", n_sites)` sites over a two-taxon tree
with branch lengths 1, the mean number of class events per site per unit
branch length should equal 1.

**Criterion.** For each scaling mode, the observed mean is within three
standard errors of 1.

```{r test1}
tree2 <- tempfile(fileext = ".newick"); writeLines("(A:1.0,B:1.0);", tree2)
p2 <- Phylogeny(tree2)

t1 <- do.call(rbind, lapply(c("synonymous-per-codon", "non-synonymous", "substitution"),
  function(sc) {
    ms <- MutationSelection(1000, 1e-8, hky, psi_1, scaling_type = sc)
    sim <- simulate_over_phylogeny(p2, ms, sample_sequence(ms, n_sites))
    sub <- sim$substitutions
    in_class <- switch(sc,
        `synonymous-per-codon` = sub$synonymous,
        "non-synonymous" = !sub$synonymous,
        substitution     = rep(TRUE, nrow(sub)))
    per_site <- tabulate(sub$site[in_class] + 1L, nbins = n_sites) / 2
    data.frame(scaling = sc, mean = mean(per_site),
               se = sd(per_site) / sqrt(n_sites),
               z = (mean(per_site) - 1) / (sd(per_site) / sqrt(n_sites)))
}))
knitr::kable(t1, digits = 4)
```

```{r test1-verdict, results="asis"}
verdict("Test 1", all(abs(t1$z) < 3),
        sprintf("class events per site per unit length: %s (|z| max %.2f, criterion < 3)",
                paste(sprintf("%.3f", t1$mean), collapse = ", "), max(abs(t1$z))))
```

# Test 2: dS calibration under homogeneous and heterogeneous models

## Homogeneous neutral mutation--selection model

The default `scaling_type = "synonymous"` is a dS gauge rather than an
event-count gauge. With amino-acid fitness differences removed, synonymous
histories divided by the model's neutral synonymous opportunity count should
recover dS = 1. The neutral total rate is also three substitutions per codon
per dS unit.

```{r test1-ds}
ms_ds <- MutationSelection(1000, 1e-8, hky, rep(0, 20),
                           scaling_type = "synonymous")
set_palantir_seed(params$seed + 1L)
sim_ds <- simulate_over_phylogeny(p2, ms_ds, sample_sequence(ms_ds, n_sites))
per_site_total <- tabulate(sim_ds$substitutions$site + 1L,
                           nbins = n_sites) / 2
per_site_syn <- tabulate(sim_ds$substitutions$site[sim_ds$substitutions$synonymous] + 1L,
                         nbins = n_sites) / 2
ds_by_site <- per_site_syn / ms_ds$synonymous_opportunities
ds_checks <- data.frame(
    quantity = c("dS from synonymous histories", "neutral events per codon"),
    expected = c(1, 3),
    observed = c(mean(ds_by_site), mean(per_site_total)),
    se = c(sd(ds_by_site), sd(per_site_total)) / sqrt(n_sites))
ds_checks$z <- (ds_checks$observed - ds_checks$expected) / ds_checks$se
knitr::kable(ds_checks, digits = 4)
```

## Heterogeneous GY94 and mutation--selection trees

**Expectation.** Under the dS gauge, branch lengths can be recovered from
complete synonymous histories after dividing by the appropriate synonymous
opportunity exposure. This should hold for (i) all branches and site classes
of a standard GY94 branch-site model and (ii) a mutation--selection branch
entered after a simultaneous fitness-profile and population-size shift. The
second expectation is computed independently by Runge--Kutta integration of
the selected master equation and the neutral dS-clock functional.

**Criterion.** Each input branch length is within four Monte-Carlo standard
errors of its history estimate. Four standard errors are used because the GY94
check evaluates six branches simultaneously.

```{r test2-ds-setup}
n_ds <- max(12000L, 4L * ceiling(n_sites / 4L))

ds_tree_file <- tempfile(fileext = ".newick")
ds_mode_file <- tempfile(fileext = ".newick")
writeLines("((A:0.15,B:0.30):0.20,(C:0.45,D:0.60):0.25);", ds_tree_file)
writeLines("((A:0,B:1):0,(C:0,D:0):0);", ds_mode_file)
ds_tree <- Phylogeny(ds_tree_file)
ds_modes <- Phylogeny(ds_mode_file, type = "mode")
ds_branches <- phylogeny_to_intervals(ds_tree, ds_modes)
ds_branches$input_dS <- ds_branches$to - ds_branches$from

set_palantir_seed(params$seed + 101L)
zny <- ZNYBranchSiteModel(
    mode_phylogeny = ds_modes,
    site_counts = rep(n_ds / 4L, 4L),
    omega0 = 0.2,
    omega2 = 3.0,
    kappa = 2.5,
    frequencies = c(T = .20, C = .30, A = .25, G = .25),
    frequency_model = "F1x4",
    scaling_type = "dS")
zny_models <- unlist(lapply(zny$models, `[[`, "models"), recursive = FALSE)
zny_NS <- vapply(zny_models, `[[`, numeric(1), "synonymous_opportunities")
stopifnot(max(zny_NS) - min(zny_NS) < 1e-12)
zny_sim <- simulate_gy94_site_model(ds_tree, zny, rescale_method = "exact")

zny_dS <- do.call(rbind, lapply(seq_len(nrow(ds_branches)), function(i) {
    node_i <- ds_branches$node[i]
    keep <- zny_sim$substitutions$node == node_i &
            zny_sim$substitutions$synonymous
    by_site <- tabulate(zny_sim$substitutions$site[keep] + 1L, nbins = n_ds)
    values <- by_site / zny_NS[1]
    estimate <- mean(values)
    se <- sd(values) / sqrt(n_ds)
    data.frame(model = "GY94 branch-site", node = node_i,
               mode = ds_branches$state[i],
               input_dS = ds_branches$input_dS[i],
               observed_dS = estimate, se = se,
               z = (estimate - ds_branches$input_dS[i]) / se)
}))
knitr::kable(zny_dS, digits = 4)
```

The mutation--selection test uses a two-tip tree so that the shifted branch is
entered directly from the ancestral equilibrium. Its independently integrated
synonymous exposure is reported with the simulation estimate.

```{r test2-mutsel-ds}
nuc <- c(T = .22, C = .28, A = .31, G = .19)
hky_ds <- HasegawaKishinoYano(nuc, transition_rate = 2.5,
                              transversion_rate = 1)
aa_order <- strsplit("ACDEFGHIKLMNPQRSTVWY", "")[[1]]
freq0 <- freq1 <- setNames(rep(.20 / 19, 20), aa_order)
freq0["W"] <- .80
freq1["A"] <- .80
Ne0 <- 500; Ne1 <- 20000
fit0 <- equilibrium_to_fitness(as.numeric(freq0), Ne0,
                               nucleotide_equilibrium = nuc)
fit1 <- equilibrium_to_fitness(as.numeric(freq1), Ne1,
                               nucleotide_equilibrium = nuc)
ms0 <- MutationSelection(Ne0, .01, hky_ds, fit0, scaling_type = "dS")
ms1 <- MutationSelection(Ne1, .01, hky_ds, fit1, scaling_type = "dS")

shift_tree_file <- tempfile(fileext = ".newick")
shift_mode_file <- tempfile(fileext = ".newick")
writeLines("(A:0.30,B:0.10);", shift_tree_file)
writeLines("(A:1,B:0);", shift_mode_file)
shift_tree <- Phylogeny(shift_tree_file)
shift_modes <- Phylogeny(shift_mode_file, type = "mode")
shift_length <- 0.30

p0_ds <- as.numeric(ms0$equilibrium)
h1_ds <- as.numeric(ms1$dS_outflux)
g1_syn <- syn_outflux(ms1)
target <- shift_length * sum(as.numeric(ms1$equilibrium) * h1_ds)
integrated <- rk4_to_budget(p0_ds, ms1$transition, h1_ds, g1_syn, target)
mean_syn_opportunity <- integrated$syn_exposure / shift_length

set_palantir_seed(params$seed + 201L)
shift_root <- sample_sequence(ms0, n_ds)
shift_sim <- simulate_over_interval_phylogeny(
    phylogeny = shift_tree,
    mode_phylogeny = shift_modes,
    models = list(ms0, ms1),
    sequence = shift_root,
    start_mode = 0,
    rescale_method = "exact")
shift_node <- shift_sim$intervals$node[shift_sim$intervals$state == 1]
shift_events <- shift_sim$substitutions[
    shift_sim$substitutions$node == shift_node &
    shift_sim$substitutions$synonymous, ]
shift_by_site <- tabulate(shift_events$site + 1L, nbins = n_ds)
shift_dS_by_site <- shift_by_site / mean_syn_opportunity
shift_estimate <- mean(shift_dS_by_site)
shift_se <- sd(shift_dS_by_site) / sqrt(n_ds)
shift_z <- (shift_estimate - shift_length) / shift_se
mutsel_dS <- data.frame(
    model = "Mutation-selection fitness + Ne shift",
    input_dS = shift_length,
    observed_dS = shift_estimate,
    se = shift_se,
    z = shift_z,
    intrinsic_time = integrated$tau,
    synonymous_opportunities_per_codon_per_dS = mean_syn_opportunity)
knitr::kable(mutsel_dS, digits = 5)
```

```{r test2-ds-verdict, results="asis"}
all_ds_z_max <- max(abs(c(ds_checks$z, zny_dS$z, shift_z)))
verdict("Test 2", all_ds_z_max < 4,
        sprintf("homogeneous checks and all history estimates recover the dS scale (|z| max %.2f, criterion < 4)",
                all_ds_z_max))
```

# Test 3: stationary distribution

**Expectation.** The model's stationary distribution (`model$equilibrium`) is
both the distribution the root sequence is sampled from and the invariant
distribution of the substitution process, so the states observed at the end
of a branch of any length are draws from it. Comparing the codon frequencies
at a tip across `r sprintf("%d", n_sites)` independent sites against the
computed stationary distribution tests the equilibrium computation, the root
sampler, and the propagation together.

**Criterion.** Monte-Carlo chi-square goodness of fit, observed tip codon
counts against the stationary distribution, with both Holm-adjusted p-values
greater than 0.01. The adjustment controls the family-wise false-positive rate
across the codon- and amino-acid-level comparisons.

```{r test3-stationary}
ms <- MutationSelection(1000, 1e-8, hky, psi_1)
sim <- simulate_over_phylogeny(p2, ms, sample_sequence(ms, n_sites))
codons <- names(GeneticCode())
obs_codon <- as.numeric(table(factor(sim$alignment["A", ], levels = codons)))
pi_codon <- as.numeric(ms$equilibrium)

chi_codon <- chisq.test(obs_codon, p = pi_codon, rescale.p = TRUE,
                        simulate.p.value = TRUE, B = 5000)
obs_aa <- codon_to_aa(obs_codon); th_aa <- codon_to_aa(pi_codon)
chi_aa <- chisq.test(obs_aa, p = th_aa, rescale.p = TRUE,
                     simulate.p.value = TRUE, B = 5000)
stationary_p <- p.adjust(c(codon = chi_codon$p.value,
                           amino_acid = chi_aa$p.value), method = "holm")
```

The stationary and observed amino-acid distributions as sequence logos
(letter height = frequency):

```{r test3-stationary-logo, fig.height=4}
plot_freqs(data.frame(stationary = th_aa, observed = obs_aa / sum(obs_aa)),
           c("stationary", "observed"))
```

```{r test3-stationary-verdict, results="asis"}
verdict("Test 3", all(stationary_p > 0.01),
        sprintf("raw chi-square p = %.3f (codon), %.3f (amino acid); minimum Holm-adjusted p = %.3f; TV(observed, stationary) = %.4f at the amino-acid level; criterion adjusted p > 0.01",
                chi_codon$p.value, chi_aa$p.value, min(stationary_p),
                tv(obs_aa / sum(obs_aa), th_aa)))
```

# Test 4: the transient between two equilibria

**Expectation.** In a time-heterogeneous simulation the sequence starts at
the stationary distribution of the first model and relaxes to the stationary
distribution of the second. The state distribution along the way follows the
master equation of the second model's generator, expressed in the branch's
own units: branch position is measured in expected substitutions of the
scaled class, so the predicted trajectory is obtained by integrating
$dp/du = pQ / g(p)$, where $g(p)$ is the class flux under the current state
distribution. This is a demanding construction: the two fitness profiles
used here are nearly disjoint, so the class flux starts at several hundred
times its equilibrium value and the branch begins with a burst of
substitutions. (This test found the defect fixed in version 1.2.1: the
exact rescaler's knot table could not resolve that burst and misplaced early
event positions.)

**Criterion.** Under both rescale methods (`"segments"` and `"exact"`):
cumulative class events per site track branch position to within
$0.02 + 4\sqrt{u/n}$ at every checkpoint (the count's sampling standard
deviation grows with $u$, so the bound scales with the checkpoint; the 1.2.1
defect exceeded the early-checkpoint bounds five-fold), and no Monte-Carlo
chi-square comparison of the observed amino-acid distribution against the
predicted trajectory is rejected at family-wise alpha 0.01 after Holm
adjustment across the twelve checkpoints.

```{r test4-transient-setup}
mA <- MutationSelection(1000, 1e-8, hky, psi_1, scaling_type = "non-synonymous")
mB <- MutationSelection(1000, 1e-8, hky, psi_2, scaling_type = "non-synonymous")
Q <- mB$transition
p0 <- as.numeric(mA$equilibrium)
piB <- as.numeric(mB$equilibrium)
W <- Q; W[outer(aa_of, aa_of, "==")] <- 0; diag(W) <- 0
w <- rowSums(W)                       # non-synonymous outflux per state

# Predicted state distribution as a function of branch position u:
# fourth-order Runge-Kutta on dp/du = pQ / g(p).
predict_traj <- function(p0, Q, w, L, du = 0.002) {
    f <- function(p) as.numeric(p %*% Q) / sum(p * w)
    n <- ceiling(L / du)
    P <- matrix(NA, n + 1, length(p0)); P[1, ] <- p0
    p <- p0
    for (k in 1:n) {
        k1 <- f(p); k2 <- f(p + du/2 * k1); k3 <- f(p + du/2 * k2); k4 <- f(p + du * k3)
        p <- p + du/6 * (k1 + 2*k2 + 2*k3 + k4)
        p[p < 0] <- 0; p <- p / sum(p)
        P[k + 1, ] <- p
    }
    list(u = seq(0, n * du, by = du), P = P)
}
L <- 8
traj <- predict_traj(p0, Q, w, L)
pred_at <- function(u) traj$P[which.min(abs(traj$u - u)), ]

treeT <- tempfile(fileext = ".newick"); writeLines(sprintf("(A:%f,B:0.01);", L), treeT)
modeT <- tempfile(fileext = ".newick"); writeLines("(A:1,B:0);", modeT)
pT <- Phylogeny(treeT); mT <- Phylogeny(modeT, type = "mode")

checkpoints <- c(0.25, 0.5, 1, 2, 4, 8)
run_method <- function(method) {
    root <- sample_sequence(mA, n_sites)
    sim <- simulate_over_interval_phylogeny(pT, mT, list(mA, mB), root,
                                            start_mode = 0, rescale_method = method)
    nodeA <- sim$intervals$node[sim$intervals$state == 1]
    subA <- sim$substitutions[sim$substitutions$node %in% nodeA, ]
    subA <- subA[order(subA$site, subA$time), ]
    ns_time <- subA$time[aa_of[subA$from + 1] != aa_of[subA$to + 1]]
    state_at <- function(u) {
        st <- root$index[, 1]
        sel <- subA[subA$time <= u, ]
        if (nrow(sel)) {
            last <- sel[!duplicated(sel$site, fromLast = TRUE), ]
            st[last$site + 1] <- last$to
        }
        st
    }
    rows <- lapply(checkpoints, function(u) {
        emp <- tabulate(state_at(u) + 1L, nbins = length(p0))
        pr <- pred_at(u)
        emp_aa <- codon_to_aa(emp); pr_aa <- codon_to_aa(pr)
        keep <- pr_aa * n_sites >= 5
        chi <- chisq.test(c(emp_aa[keep], sum(emp_aa[!keep])),
                          p = c(pr_aa[keep], sum(pr_aa[!keep])),
                          rescale.p = TRUE, simulate.p.value = TRUE, B = 5000)
        data.frame(method = method, u = u,
                   cum_class = sum(ns_time <= u) / n_sites,
                   p = chi$p.value,
                   tv_obs_pred = tv(emp_aa / n_sites, pr_aa),
                   tv_pred_eqB = tv(pr_aa, codon_to_aa(piB)))
    })
    list(table = do.call(rbind, rows), state_at = state_at)
}
res_seg <- run_method("segments")
res_ex  <- run_method("exact")
t4 <- rbind(res_seg$table, res_ex$table)
t4$p_holm <- p.adjust(t4$p, method = "holm")
knitr::kable(t4, digits = 4)
```

The distance from the new equilibrium decays along the branch, and the
observed distributions (points) follow the predicted trajectory (line) under
both methods:

```{r test4-transient-curve}
u_dense <- seq(0, L, by = 0.1)
tv_pred <- sapply(u_dense, function(u) tv(codon_to_aa(pred_at(u)), codon_to_aa(piB)))
plot(u_dense, tv_pred, type = "l", lwd = 2,
     xlab = "Branch position (expected non-synonymous substitutions per site)",
     ylab = "Total variation distance from the new equilibrium")
u_pts <- seq(0.1, L, by = 0.26)
for (r in list(list(res_seg, "#E84B2A", 19), list(res_ex, "#2166AC", 1))) {
    tv_obs <- sapply(u_pts, function(u)
        tv(codon_to_aa(tabulate(r[[1]]$state_at(u) + 1L, nbins = length(p0))) / n_sites,
           codon_to_aa(piB)))
    points(u_pts, tv_obs, col = r[[2]], pch = r[[3]])
}
legend("topright", bty = "n",
       c("predicted trajectory", "observed (segments)", "observed (exact)"),
       lwd = c(2, NA, NA), pch = c(NA, 19, 1), col = c("black", "#E84B2A", "#2166AC"))
```

Amino-acid distributions at the start, in mid-transient, and at the end of
the branch — predicted against observed (segments method):

```{r test4-transient-logo, fig.height=4}
plot_freqs(data.frame(
    start_pred = codon_to_aa(p0),
    u0.5_pred  = codon_to_aa(pred_at(0.5)),
    u0.5_obs   = codon_to_aa(tabulate(res_seg$state_at(0.5) + 1L, nbins = length(p0))) / n_sites,
    u8_pred    = codon_to_aa(pred_at(8)),
    u8_obs     = codon_to_aa(tabulate(res_seg$state_at(8) + 1L, nbins = length(p0))) / n_sites,
    eq_B       = codon_to_aa(piB)),
    c("start", "u=0.5 pred", "u=0.5 obs", "u=8 pred", "u=8 obs", "equilibrium B"))
```

```{r test4-transient-verdict, results="asis"}
calib_tol <- 0.02 + 4 * sqrt(t4$u / n_sites)
calib_ok <- all(abs(t4$cum_class - t4$u) < calib_tol)
p_ok <- all(t4$p_holm > 0.01)
verdict("Test 4", calib_ok && p_ok,
        sprintf("max |cumulative class events - u| / tolerance = %.2f (criterion < 1); raw chi-square p range %.3f-%.3f, minimum Holm-adjusted p %.3f (criterion > 0.01)",
                max(abs(t4$cum_class - t4$u) / calib_tol), min(t4$p),
                max(t4$p), min(t4$p_holm)))
```

# Summary

```{r summary, results="asis"}
for (id in names(verdicts)) {
    cat(sprintf("- **%s: %s**\n", id, ifelse(verdicts[[id]], "PASS", "FAIL")))
}
cat(sprintf("\n**Overall: %s** (PalantiR %s, seed %d, %d sites in standard tests; %d in tree-level dS tests)\n",
    ifelse(all(unlist(verdicts)), "PASS", "FAIL"),
    as.character(packageVersion("PalantiR")), params$seed, n_sites, n_ds))
```
