---
title: "PalantiR engine validation"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
params:
  seed: 20260821
  n_sites: 3000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, warning = FALSE, message = FALSE)
suppressMessages(library(PalantiR))
set_palantir_seed(params$seed)

verdicts <- list()
verdict <- function(id, ok, detail) {
    verdicts[[id]] <<- ok
    cat(sprintf("**%s: %s** — %s\n", id, ifelse(ok, "PASS", "FAIL"), detail))
}
has_logo <- requireNamespace("ggseqlogo", quietly = TRUE)

aa_of <- unname(GeneticCode())
aa_levels <- sort(unique(aa_of))

# Sequence logo of one or more amino-acid frequency distributions
# (columns of `freqs`), with a barplot fallback if ggseqlogo is absent.
plot_freqs <- function(freqs, labels) {
    if (has_logo) {
        m <- as.matrix(freqs); rownames(m) <- aa_levels; colnames(m) <- labels
        print(ggseqlogo::ggseqlogo(m, method = "custom", seq_type = "aa") +
              ggplot2::scale_x_continuous(breaks = seq_along(labels), labels = labels) +
              ggplot2::ylab("frequency"))
    } else {
        barplot(t(as.matrix(freqs)), beside = TRUE, names.arg = aa_levels,
                legend.text = labels, ylab = "frequency")
    }
}

codon_to_aa <- function(codon_counts) as.numeric(tapply(codon_counts, aa_of, sum))
tv <- function(p, q) 0.5 * sum(abs(p - q))
```

This notebook tests the core simulation engine against exact expectations.
Each test states its expectation, runs a fresh simulation, compares the
result statistically and visually, and prints a PASS or FAIL verdict against
a criterion stated in advance. To rerun everything with a different seed or
simulation size:

```r
rmarkdown::render("engine_validation.Rmd", params = list(seed = 1, n_sites = 3000))
```

All tests use the fitness profiles shipped with the package (`psi_c50_1`,
amino-acid equilibrium frequencies converted to fitness with
`equilibrium_to_fitness`) and the mutation–selection codon model.

```{r inputs}
data(psi_c50_1)
hky <- HasegawaKishinoYano(equilibrium = rep(0.25, 4))
psi_1 <- equilibrium_to_fitness(as.numeric(psi_c50_1[1, ]), 1000)
psi_2 <- equilibrium_to_fitness(as.numeric(psi_c50_1[2, ]), 1000)
n_sites <- params$n_sites
```

# Test 1: branch-length calibration under the three scaling modes

**Expectation.** A model built with `scaling_type = "synonymous"`,
`"non-synonymous"`, or `"substitution"` is normalized so that one unit of
branch length delivers, at stationarity, one expected substitution of that
class per codon site: synonymous, non-synonymous, or of any kind,
respectively. Simulating `r sprintf("%d", 3000)` sites over a two-taxon tree
with branch lengths 1, the mean number of class events per site per unit
branch length should equal 1.

**Criterion.** For each scaling mode, the observed mean is within three
standard errors of 1.

```{r test1}
tree2 <- tempfile(fileext = ".newick"); writeLines("(A:1.0,B:1.0);", tree2)
p2 <- Phylogeny(tree2)

t1 <- do.call(rbind, lapply(c("synonymous", "non-synonymous", "substitution"),
  function(sc) {
    ms <- MutationSelection(1000, 1e-8, hky, psi_1, scaling_type = sc)
    sim <- simulate_over_phylogeny(p2, ms, sample_sequence(ms, n_sites))
    sub <- sim$substitutions
    in_class <- switch(sc,
        synonymous       = sub$synonymous,
        "non-synonymous" = !sub$synonymous,
        substitution     = rep(TRUE, nrow(sub)))
    per_site <- tabulate(sub$site[in_class] + 1L, nbins = n_sites) / 2
    data.frame(scaling = sc, mean = mean(per_site),
               se = sd(per_site) / sqrt(n_sites),
               z = (mean(per_site) - 1) / (sd(per_site) / sqrt(n_sites)))
}))
knitr::kable(t1, digits = 4)
```

```{r test1-verdict, results="asis"}
verdict("Test 1", all(abs(t1$z) < 3),
        sprintf("class events per site per unit length: %s (|z| max %.2f, criterion < 3)",
                paste(sprintf("%.3f", t1$mean), collapse = ", "), max(abs(t1$z))))
```

# Test 2: stationary distribution

**Expectation.** The model's stationary distribution (`model$equilibrium`) is
both the distribution the root sequence is sampled from and the invariant
distribution of the substitution process, so the states observed at the end
of a branch of any length are draws from it. Comparing the codon frequencies
at a tip across `r sprintf("%d", 3000)` independent sites against the
computed stationary distribution tests the equilibrium computation, the root
sampler, and the propagation together.

**Criterion.** Monte-Carlo chi-square goodness of fit, observed tip codon
counts against the stationary distribution, p > 0.01 at both the codon and
the amino-acid level.

```{r test2}
ms <- MutationSelection(1000, 1e-8, hky, psi_1)
sim <- simulate_over_phylogeny(p2, ms, sample_sequence(ms, n_sites))
codons <- names(GeneticCode())
obs_codon <- as.numeric(table(factor(sim$alignment["A", ], levels = codons)))
pi_codon <- as.numeric(ms$equilibrium)

chi_codon <- chisq.test(obs_codon, p = pi_codon, rescale.p = TRUE,
                        simulate.p.value = TRUE, B = 5000)
obs_aa <- codon_to_aa(obs_codon); th_aa <- codon_to_aa(pi_codon)
chi_aa <- chisq.test(obs_aa, p = th_aa, rescale.p = TRUE,
                     simulate.p.value = TRUE, B = 5000)
```

The stationary and observed amino-acid distributions as sequence logos
(letter height = frequency):

```{r test2-logo, fig.height=4}
plot_freqs(data.frame(stationary = th_aa, observed = obs_aa / sum(obs_aa)),
           c("stationary", "observed"))
```

```{r test2-verdict, results="asis"}
verdict("Test 2", chi_codon$p.value > 0.01 && chi_aa$p.value > 0.01,
        sprintf("chi-square p = %.3f (codon), %.3f (amino acid); TV(observed, stationary) = %.4f at the amino-acid level; criterion p > 0.01",
                chi_codon$p.value, chi_aa$p.value, tv(obs_aa / sum(obs_aa), th_aa)))
```

# Test 3: the transient between two equilibria

**Expectation.** In a time-heterogeneous simulation the sequence starts at
the stationary distribution of the first model and relaxes to the stationary
distribution of the second. The state distribution along the way follows the
master equation of the second model's generator, expressed in the branch's
own units: branch position is measured in expected substitutions of the
scaled class, so the predicted trajectory is obtained by integrating
$dp/du = pQ / g(p)$, where $g(p)$ is the class flux under the current state
distribution. This is a demanding construction: the two fitness profiles
used here are nearly disjoint, so the class flux starts at several hundred
times its equilibrium value and the branch begins with a burst of
substitutions. (This test found the defect fixed in version 1.2.1: the
exact rescaler's knot table could not resolve that burst and misplaced early
event positions.)

**Criterion.** Under both rescale methods (`"segments"` and `"exact"`):
cumulative class events per site track branch position to within 0.04 at
every checkpoint, and Monte-Carlo chi-square of the observed amino-acid
distribution against the predicted trajectory gives p ≥ 0.001 at every
checkpoint with at most two of the twelve p-values below 0.05.

```{r test3-setup}
mA <- MutationSelection(1000, 1e-8, hky, psi_1, scaling_type = "non-synonymous")
mB <- MutationSelection(1000, 1e-8, hky, psi_2, scaling_type = "non-synonymous")
Q <- mB$transition
p0 <- as.numeric(mA$equilibrium)
piB <- as.numeric(mB$equilibrium)
W <- Q; W[outer(aa_of, aa_of, "==")] <- 0; diag(W) <- 0
w <- rowSums(W)                       # non-synonymous outflux per state

# Predicted state distribution as a function of branch position u:
# fourth-order Runge-Kutta on dp/du = pQ / g(p).
predict_traj <- function(p0, Q, w, L, du = 0.002) {
    f <- function(p) as.numeric(p %*% Q) / sum(p * w)
    n <- ceiling(L / du)
    P <- matrix(NA, n + 1, length(p0)); P[1, ] <- p0
    p <- p0
    for (k in 1:n) {
        k1 <- f(p); k2 <- f(p + du/2 * k1); k3 <- f(p + du/2 * k2); k4 <- f(p + du * k3)
        p <- p + du/6 * (k1 + 2*k2 + 2*k3 + k4)
        p[p < 0] <- 0; p <- p / sum(p)
        P[k + 1, ] <- p
    }
    list(u = seq(0, n * du, by = du), P = P)
}
L <- 8
traj <- predict_traj(p0, Q, w, L)
pred_at <- function(u) traj$P[which.min(abs(traj$u - u)), ]

treeT <- tempfile(fileext = ".newick"); writeLines(sprintf("(A:%f,B:0.01);", L), treeT)
modeT <- tempfile(fileext = ".newick"); writeLines("(A:1,B:0);", modeT)
pT <- Phylogeny(treeT); mT <- Phylogeny(modeT, type = "mode")

checkpoints <- c(0.25, 0.5, 1, 2, 4, 8)
run_method <- function(method) {
    root <- sample_sequence(mA, n_sites)
    sim <- simulate_over_interval_phylogeny(pT, mT, list(mA, mB), root,
                                            start_mode = 0, rescale_method = method)
    nodeA <- sim$intervals$node[sim$intervals$state == 1]
    subA <- sim$substitutions[sim$substitutions$node %in% nodeA, ]
    subA <- subA[order(subA$site, subA$time), ]
    ns_time <- subA$time[aa_of[subA$from + 1] != aa_of[subA$to + 1]]
    state_at <- function(u) {
        st <- root$index[, 1]
        sel <- subA[subA$time <= u, ]
        if (nrow(sel)) {
            last <- sel[!duplicated(sel$site, fromLast = TRUE), ]
            st[last$site + 1] <- last$to
        }
        st
    }
    rows <- lapply(checkpoints, function(u) {
        emp <- tabulate(state_at(u) + 1L, nbins = length(p0))
        pr <- pred_at(u)
        emp_aa <- codon_to_aa(emp); pr_aa <- codon_to_aa(pr)
        keep <- pr_aa * n_sites >= 5
        chi <- chisq.test(c(emp_aa[keep], sum(emp_aa[!keep])),
                          p = c(pr_aa[keep], sum(pr_aa[!keep])),
                          rescale.p = TRUE, simulate.p.value = TRUE, B = 5000)
        data.frame(method = method, u = u,
                   cum_class = sum(ns_time <= u) / n_sites,
                   p = chi$p.value,
                   tv_obs_pred = tv(emp_aa / n_sites, pr_aa),
                   tv_pred_eqB = tv(pr_aa, codon_to_aa(piB)))
    })
    list(table = do.call(rbind, rows), state_at = state_at)
}
res_seg <- run_method("segments")
res_ex  <- run_method("exact")
t3 <- rbind(res_seg$table, res_ex$table)
knitr::kable(t3, digits = 4)
```

The distance from the new equilibrium decays along the branch, and the
observed distributions (points) follow the predicted trajectory (line) under
both methods:

```{r test3-curve}
u_dense <- seq(0, L, by = 0.1)
tv_pred <- sapply(u_dense, function(u) tv(codon_to_aa(pred_at(u)), codon_to_aa(piB)))
plot(u_dense, tv_pred, type = "l", lwd = 2,
     xlab = "Branch position (expected non-synonymous substitutions per site)",
     ylab = "Total variation distance from the new equilibrium")
u_pts <- seq(0.1, L, by = 0.26)
for (r in list(list(res_seg, "#E84B2A", 19), list(res_ex, "#2166AC", 1))) {
    tv_obs <- sapply(u_pts, function(u)
        tv(codon_to_aa(tabulate(r[[1]]$state_at(u) + 1L, nbins = length(p0))) / n_sites,
           codon_to_aa(piB)))
    points(u_pts, tv_obs, col = r[[2]], pch = r[[3]])
}
legend("topright", bty = "n",
       c("predicted trajectory", "observed (segments)", "observed (exact)"),
       lwd = c(2, NA, NA), pch = c(NA, 19, 1), col = c("black", "#E84B2A", "#2166AC"))
```

Amino-acid distributions at the start, in mid-transient, and at the end of
the branch — predicted against observed (segments method):

```{r test3-logo, fig.height=4}
plot_freqs(data.frame(
    start_pred = codon_to_aa(p0),
    u0.5_pred  = codon_to_aa(pred_at(0.5)),
    u0.5_obs   = codon_to_aa(tabulate(res_seg$state_at(0.5) + 1L, nbins = length(p0))) / n_sites,
    u8_pred    = codon_to_aa(pred_at(8)),
    u8_obs     = codon_to_aa(tabulate(res_seg$state_at(8) + 1L, nbins = length(p0))) / n_sites,
    eq_B       = codon_to_aa(piB)),
    c("start", "u=0.5 pred", "u=0.5 obs", "u=8 pred", "u=8 obs", "equilibrium B"))
```

```{r test3-verdict, results="asis"}
calib_ok <- all(abs(t3$cum_class - t3$u) < 0.04)
p_ok <- min(t3$p) >= 0.001 && sum(t3$p < 0.05) <= 2
verdict("Test 3", calib_ok && p_ok,
        sprintf("max |cumulative class events - u| = %.3f (criterion < 0.04); chi-square p range %.3f-%.3f with %d of %d below 0.05 (criterion: min >= 0.001, at most 2)",
                max(abs(t3$cum_class - t3$u)), min(t3$p), max(t3$p), sum(t3$p < 0.05), nrow(t3)))
```

# Summary

```{r summary, results="asis"}
for (id in names(verdicts)) {
    cat(sprintf("- **%s: %s**\n", id, ifelse(verdicts[[id]], "PASS", "FAIL")))
}
cat(sprintf("\n**Overall: %s** (PalantiR %s, seed %d, %d sites per test)\n",
    ifelse(all(unlist(verdicts)), "PASS", "FAIL"),
    as.character(packageVersion("PalantiR")), params$seed, n_sites))
```
