---
output:
  html_document:
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
set_palantir_seed(20260820)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Branch Heterogeneity

Branch heterogeneous simulations specify different models (called `modes`) for each branch on a phylogeny. 
This is done by providing a newick tree with model indexes instead of branch length, and a list of models.

For more details on time-heterogeneous simulation, see [Temporal Heterogeneity Rescaling](Temporal_Heterogeneity_Rescaling.html).

Here is an example file, which represents a model switch on the primate lineage in the mammalian phylogeny:
```{r}
scan("../inst/extdata/mammals_switch.newick", what = character(), quiet = T)
```

This phylogeny can be loaded with the `Phylogeny` constructor, with an additional `type = "mode"` argument.

```{r}
# Load phylogeny
mammals <- Phylogeny("../inst/extdata/mammals.newick")

# Load branch modes
mammal_switches <- Phylogeny("../inst/extdata/mammals_switch.newick", type = "mode")
```

We can view the phylogeny and the assigned modes:

```{r}
plot(mammals, mammal_switches)
```

# Example

In the example below, we simulate on a mammalian phylogeny with a population size expansion on the primate lineage.

## Setup

Load model parameters:

```{r}
# Load nucleotide equilibrium distribution
# Nucleotide order is T,C,A,G
nuc_pi <- c(.3, .2, .25, .25)

# Load amino acid equilibrium frequencies
# These will be converted to population-specific fitness vectors
# (the raw file carries a spurious 21st column from a trailing tab,
# so keep the twenty amino-acid columns)
aa_pi <- as.matrix(read.table("../inst/extdata/psi-c50-1.txt"))[, 1:20]

psi <- equilibrium_to_fitness(aa_pi[1, ], 5000)
```

Create models:

```{r}
# Create nucleotide substitution model
hky <- HasegawaKishinoYano(nuc_pi)

# Create codon models

# Smaller population size
ms_5k <- MutationSelection(
    population_size = 5000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

# Larger population size
ms_30k <- MutationSelection(
    population_size = 30000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

```

Sample sequence from model equilibrium distribution to start at root:
```{r sample_sequence}
s <- sample_sequence(ms_5k, 100)
```

Simulate:
```{r simulate}
sim <- simulate_over_interval_phylogeny(
    phylogeny = mammals,
    mode_phylogeny = mammal_switches,
    models = list(ms_5k, ms_30k),
    sequence = s,
    start_mode = 0)
```

Note that we have to specify which mode the root starts with using `start_mode` argument.

## Substitution history

Plot substitution history:

```{r plot_sim}
plot(sim, sites = 0:24, circle_opacity = 0.6)
```

Synonymous substitutions are shown in green, non-synonymous in red, and branch modes as colored
branches. An increase in population size is expected to produce a transient excess of non-synonymous
substitutions, as selection becomes strong enough to replace mildly deleterious variants that fixed
under the smaller population size. The shifted clade is short, so all of its branches fall within this
transient period.

The two regimes differ more in the direction of amino-acid change than in its rate. At stationarity,
substitutions that increase fitness and substitutions that decrease it occur at equal rates, so the mean
fitness change per substitution is zero. After the expansion, fitness-increasing substitutions
predominate until the clade reaches its new equilibrium. The `intervals` table assigns each branch to
its model, which allows the shifted clade to be compared with the remainder of the tree:

```{r, fig.height=4.5}
aa <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L",
        "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y")
nsyn <- subset(sim$substitutions, !synonymous)
nsyn$group <- ifelse(sim$intervals$state[match(nsyn$node, sim$intervals$node)] == 1,
                     "shifted clade", "rest of tree")
d_psi <- psi[match(nsyn$amino_acid_to, aa)] - psi[match(nsyn$amino_acid_from, aa)]
s_scaled <- 2 * 5000 * d_psi   # scaled by the ancestral population size

counts <- rbind(gain = tapply(d_psi > 0, nsyn$group, sum),
                loss = tapply(d_psi < 0, nsyn$group, sum))
mean_s <- tapply(s_scaled, nsyn$group, mean)
se_s   <- tapply(s_scaled, nsyn$group, function(x) sd(x) / sqrt(length(x)))

par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 1))
bars <- barplot(counts, beside = TRUE, col = c("#2E8B57", "#C0392B"),
                ylab = "Non-synonymous substitutions", ylim = c(0, max(counts) * 1.15),
                legend.text = c("fitness gain", "fitness loss"),
                args.legend = list(bty = "n", x = "topright"))
text(bars, counts, labels = counts, pos = 3, cex = 0.9)
bars2 <- barplot(mean_s, col = "#8DA0CB", ylab = "Mean scaled fitness change (2Ns)",
                 ylim = range(0, mean_s + sign(mean_s) * (se_s + 0.3)))
arrows(bars2, mean_s - se_s, bars2, mean_s + se_s, angle = 90, code = 3, length = 0.05)
abline(h = 0)
```

In the rest of the tree, gains and losses occur in nearly equal numbers and the mean population-scaled
fitness change is close to zero, as expected at stationarity. In the shifted clade, gains outnumber
losses 104 to 5 and the mean scaled change per substitution is +1.4 (error bars are standard errors of
the mean).

Check substitution history table:

```{r, results='hide'}
head(sim$substitutions)
```
```{r, echo=F, results='asis'}
kable(head(sim$substitutions))
```

## Alignment

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```

Show alignment in a scrollable viewer:

```{r}
plot(sim$alignment)
```
