---
output:
  html_document:
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
set_palantir_seed(20260820)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Branch Heterogeneity

Branch heterogeneous simulations specify different models (called `modes`) for each branch on a phylogeny. 
This is done by providing a newick tree with model indexes instead of branch length, and a list of models.

For more details on time-heterogeneous simulation, see [Temporal Heterogeneity Rescaling](Temporal_Heterogeneity_Rescaling.html).

Here is an example file, which represents a model switch on the primate lineage in the mammalian phylogeny:
```{r}
scan("../inst/extdata/mammals_switch.newick", what = character(), quiet = T)
```

This phylogeny can be loaded with the `Phylogeny` constructor, with an additional `type = "mode"` argument.

```{r}
# Load phylogeny
mammals <- Phylogeny("../inst/extdata/mammals.newick")

# Load branch modes
mammal_switches <- Phylogeny("../inst/extdata/mammals_switch.newick", type = "mode")
```

We can view the phylogeny and the assigned modes:

```{r}
plot(mammals, mammal_switches)
```

# Example

In the example below, we simulate on a mammalian phylogeny with a population size expansion on the primate lineage.

## Setup

Load model parameters:

```{r}
# Load nucleotide equilibrium distribution
# Nucleotide order is T,C,A,G
nuc_pi <- c(.3, .2, .25, .25)

# Load amino acid equilibrium frequencies
# These will be converted to population-specific fitness vectors
# (the raw file carries a spurious 21st column from a trailing tab,
# so keep the twenty amino-acid columns)
aa_pi <- as.matrix(read.table("../inst/extdata/psi-c50-1.txt"))[, 1:20]

psi <- equilibrium_to_fitness(aa_pi[1, ], 5000)
```

Create models:

```{r}
# Create nucleotide substitution model
hky <- HasegawaKishinoYano(nuc_pi)

# Create codon models

# Smaller population size
ms_5k <- MutationSelection(
    population_size = 5000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

# Larger population size
ms_8k <- MutationSelection(
    population_size = 8000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

```

Sample sequence from model equilibrium distribution to start at root:
```{r sample_sequence}
s <- sample_sequence(ms_5k, 100)
```

Simulate:
```{r simulate}
sim <- simulate_over_interval_phylogeny(
    phylogeny = mammals,
    mode_phylogeny = mammal_switches,
    models = list(ms_5k, ms_8k),
    sequence = s,
    start_mode = 0)
```

Note that we have to specify which mode the root starts with using `start_mode` argument.

## Substitution history

Plot substitution history:

```{r plot_sim}
plot(sim, sites = 0:24, circle_opacity = 0.6)
```

Synonymous substitutions are shown in green, non-synonymous in red, and branch modes as colored branches.
A population-size increase is expected to produce a transient burst of non-synonymous substitutions just after
the shift -- stronger selection removes variants that had drifted to fixation under the smaller population --
followed by a slightly lower long-run non-synonymous rate. With a modest expansion like this one (5,000 to
8,000) the signature is subtle, and the two regimes look broadly similar in a single 100-site simulation: a
realistic picture of how weak the footprint of a moderate population-size shift is.

Check substitution history table:

```{r, results='hide'}
head(sim$substitutions)
```
```{r, echo=F, results='asis'}
kable(head(sim$substitutions))
```

## Alignment

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```

Show alignment in a scrollable viewer:

```{r}
plot(sim$alignment)
```
