---
output:
  html_document:
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
set_palantir_seed(20260820)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Branch Heterogeneity

Branch heterogeneous simulations specify different models (called `modes`) for each branch on a phylogeny. 
This is done by providing a newick tree with model indexes instead of branch length, and a list of models.

For more details on time-heterogeneous simulation, see [Temporal Heterogeneity Rescaling](Temporal_Heterogeneity_Rescaling.html).

Here is an example file, which represents a model switch on the primate lineage in the mammalian phylogeny:
```{r}
scan("../inst/extdata/mammals_switch.newick", what = character(), quiet = T)
```

This phylogeny can be loaded with the `Phylogeny` constructor, with an additional `type = "mode"` argument.

```{r}
# Load phylogeny
mammals <- Phylogeny("../inst/extdata/mammals.newick")

# Load branch modes
mammal_switches <- Phylogeny("../inst/extdata/mammals_switch.newick", type = "mode")
```

We can view the phylogeny and the assigned modes:

```{r}
plot(mammals, mammal_switches)
```

# Example

In the example below, we simulate on a mammalian phylogeny with a population size expansion on the primate lineage.

## Setup

Load model parameters:

```{r}
# Load nucleotide equilibrium distribution
# Nucleotide order is T,C,A,G
nuc_pi <- c(.3, .2, .25, .25)

# Load amino acid equilibrium frequencies
# These will be converted to population-specific fitness vectors
# (the raw file carries a spurious 21st column from a trailing tab,
# so keep the twenty amino-acid columns)
aa_pi <- as.matrix(read.table("../inst/extdata/psi-c50-1.txt"))[, 1:20]

psi <- equilibrium_to_fitness(aa_pi[1, ], 5000)
```

Create models:

```{r}
# Create nucleotide substitution model
hky <- HasegawaKishinoYano(nuc_pi)

# Create codon models

# Smaller population size
ms_5k <- MutationSelection(
    population_size = 5000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

# Larger population size
ms_30k <- MutationSelection(
    population_size = 30000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = psi)

```

Sample sequence from model equilibrium distribution to start at root:
```{r sample_sequence}
s <- sample_sequence(ms_5k, 100)
```

Simulate:
```{r simulate}
sim <- simulate_over_interval_phylogeny(
    phylogeny = mammals,
    mode_phylogeny = mammal_switches,
    models = list(ms_5k, ms_30k),
    sequence = s,
    start_mode = 0)
```

Note that we have to specify which mode the root starts with using `start_mode` argument.

## Substitution history

Plot substitution history:

```{r plot_sim}
plot(sim, sites = 0:24, circle_opacity = 0.6)
```

Synonymous substitutions are shown in green, non-synonymous in red, and branch modes as colored branches.
A population-size increase produces a transient burst of non-synonymous substitutions after the shift --
stronger selection removes variants that had drifted to fixation under the smaller population -- before
settling to a lower long-run rate. The shifted clade here is short enough that it sits inside that
transient: its non-synonymous rate is modestly elevated, and the fitness direction of its changes is
strongly biased.

The direction bias is the clearest signature. At stationarity, uphill and downhill substitutions balance
exactly; after the expansion, purging makes uphill changes dominate until the clade re-equilibrates. The
`intervals` table maps each branch to its mode, so the comparison is direct:

```{r}
aa <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L",
        "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y")
nsyn <- subset(sim$substitutions, !synonymous)
nsyn$mode <- sim$intervals$state[match(nsyn$node, sim$intervals$node)]
d_psi <- psi[match(nsyn$amino_acid_to, aa)] - psi[match(nsyn$amino_acid_from, aa)]
direction <- ifelse(d_psi > 0, "uphill", ifelse(d_psi < 0, "downhill", "level"))
table(mode = nsyn$mode, direction)
```

The constant-population branches sit near the stationary uphill:downhill balance, while on the
expanded-population branches nearly every amino-acid change is an uphill move.

## Visualizing the purging transient

The synonymous/non-synonymous colouring above cannot show this bias, but the plot honours per-event
colours through the substitution table's `color` column. Recolouring each substitution by its fitness
direction makes the purging visible on the tree itself:

```{r}
sub <- sim$substitutions
d_all <- psi[match(sub$amino_acid_to, aa)] - psi[match(sub$amino_acid_from, aa)]
sub$color <- ifelse(sub$synonymous, "#BBBBBB",
             ifelse(d_all > 0, "#E84B2A", ifelse(d_all < 0, "#2166AC", "#888888")))
sim_dir <- sim
sim_dir$substitutions <- sub
plot(sim_dir, sites = 0:24, circle_opacity = 0.6)
```

Uphill changes are red, downhill blue, synonymous light grey. Outside the shifted clade red and blue
appear in near-equal numbers; inside it the substitutions are almost exclusively red.

To see that the burst is concentrated at the base of the shifted clade, measure the substitution rate as
a function of time since the shift, normalising each time slice by the lineage length present in it. The
branch start depths come from a short walk over the phylogeny's JSON representation:

```{r}
library(jsonlite)
branches <- data.frame()
walk <- function(node, start) {
  len <- if (is.null(node$length)) 0 else node$length
  branches <<- rbind(branches, data.frame(node = node$index, start = start, length = len))
  if (!is.null(node$children)) for (ch in node$children) walk(ch, start + len)
}
walk(fromJSON(mammals$json, simplifyVector = FALSE), 0)
branches$mode <- sim$intervals$state[match(branches$node, sim$intervals$node)]

switch_depth <- min(branches$start[branches$mode == 1], na.rm = TRUE)
nsyn$depth <- branches$start[match(nsyn$node, branches$node)] + nsyn$time - switch_depth

clade <- subset(branches, mode == 1 & length > 0)
clade$s0 <- clade$start - switch_depth
edges <- seq(0, max(clade$s0 + clade$length) + 1e-9, length.out = 8)
exposure <- sapply(seq_len(length(edges) - 1), function(i)
  100 * sum(pmax(0, pmin(clade$s0 + clade$length, edges[i + 1]) - pmax(clade$s0, edges[i]))))
up <- hist(nsyn$depth[nsyn$mode == 1 & direction == "uphill"],   breaks = edges, plot = FALSE)$counts
dn <- hist(nsyn$depth[nsyn$mode == 1 & direction == "downhill"], breaks = edges, plot = FALSE)$counts

total_m0 <- 100 * sum(branches$length[branches$mode == 0], na.rm = TRUE)
base_up <- sum(nsyn$mode == 0 & direction == "uphill",   na.rm = TRUE) / total_m0
base_dn <- sum(nsyn$mode == 0 & direction == "downhill", na.rm = TRUE) / total_m0
```

```{r, fig.height=5.5}
mids <- (head(edges, -1) + tail(edges, -1)) / 2
up_rate <- up / exposure
se <- sqrt(up) / exposure
plot(mids, up_rate, type = "b", pch = 19, col = "#E84B2A",
     ylim = c(0, max(up_rate + se) * 1.05),
     xlab = "Time since the population-size shift (branch-length units)",
     ylab = "Non-synonymous substitutions / site / unit length")
arrows(mids, up_rate - se, mids, up_rate + se,
       angle = 90, code = 3, length = 0.04, col = "#E84B2A")
lines(mids, dn / exposure, type = "b", pch = 19, col = "#2166AC")
abline(h = base_up, lty = 2, col = "#E84B2A")
abline(h = base_dn, lty = 2, col = "#2166AC")
legend("topright", bty = "n",
       c("uphill, shifted clade", "downhill, shifted clade",
         "uphill baseline (constant N)", "downhill baseline (constant N)"),
       col = rep(c("#E84B2A", "#2166AC"), 2), lty = c(1, 1, 2, 2), pch = c(19, 19, NA, NA))
```

The uphill rate immediately after the shift is about four times the constant-N baseline and roughly
double its own later level, while the downhill rate collapses to near zero. Even deep in the clade the
uphill rate stays about twice the baseline: the purging transient is strongest at the stem and the clade
has not yet re-equilibrated to the larger population size (error bars are Poisson standard errors from a
single 100-site simulation).

Check substitution history table:

```{r, results='hide'}
head(sim$substitutions)
```
```{r, echo=F, results='asis'}
kable(head(sim$substitutions))
```

## Alignment

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```

Show alignment in a scrollable viewer:

```{r}
plot(sim$alignment)
```
