---
output:
  html_document:
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
set_palantir_seed(20260820)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Co-Evolution

Co-Evolution is an extension of a Mutation-Selection codon model, which considers pairs of codons.

The model adds an amino-acid correlation matrix $\delta$, which expresses how "linked" two amino acid states are within a pair. 

For example, $\delta_{A, K} > 1$ means that $AK$ is a favoured pair, while $\delta_{W, Y} < 1$ means $WY$ is disfavoured. $\delta = 1$ signifies lack of association.

The selection coefficient is calculated for a pair of sites, assuming additive fitness.

For a substitution from codon pair $ij$ to codon pair $kj$ (a change at the
first site; changes at the second site are symmetric):

$$
s_{ij, kj} = (\psi_k + \psi_j)\delta_{kj} - (\psi_i + \psi_j)\delta_{ij} \\
Q_{ij, kj} = 2N_e \, \mu_{ik} \, \frac{1-e ^ {-s_{ij, kj}}}{1-e^{-2 N_e s_{ij, kj}}}
$$

Only single-nucleotide codon changes at a single site of the pair have
non-zero instantaneous rate; simultaneous changes at both sites have rate
zero (as of version 1.2.0), so compensated pairs arise as two rapid
sequential substitutions.

* $N_e$ is the effective population size
* $\mu_{ij}$ is the mutation rate between a pair of nucleotides. This rate is taken from a nucleotide substitution model, such as Hasegawa Kishino Yano [[1](https://www.ncbi.nlm.nih.gov/pubmed/3934395)].
* $\psi_i$ is the absolute fitness of amino acid for codon $i$. Values of $\psi$ for the first and second pair can be specified from different fitness vectors.
* $s_{ij, kl}$ is selection coefficient for a pair substitution. If either substitution in a pair is synonymous, the second term of the product will be equal to $1$.

# Example: compensatory evolution at a salt bridge

We construct a pair landscape in which only two amino-acid configurations
have high fitness: lysine--glutamate and glutamate--lysine, the classic
charge-swap of an electrostatic interaction. The single-mismatch
configurations (K,K and E,E) are deleterious but passable, and every other
configuration is strongly deleterious. With this landscape the pair toggles
between K,E and E,K through short-lived intermediates: a substitution at
one site followed almost immediately by the compensating substitution at
the other.

## Setup

```{r}
# Load phylogeny
mammals <- Phylogeny("../inst/extdata/mammals.newick")

# Flat site fitness: all structure comes from the pair coupling
f <- rep(1, 20)
Ne <- 1000

# Amino acid order is:
# A, C, D, E, F, G, H, I, K, L, M, N, P, Q, R, S, T, V, W, Y
K <- 9; E <- 4

# Population-scaled selection coefficients (2*Ne*s) of each pair
# configuration, relative to the compatible charge-swap pairs:
S <- matrix(-20, 20, 20)   # everything else:  2Ns = -20
S[K, K] <- -8              # mismatch K,K:     2Ns =  -8
S[E, E] <- -8              # mismatch E,E:     2Ns =  -8
S[K, E] <- 0               # compatible:       2Ns =   0
S[E, K] <- 0

# The scaled fitness of pair (i, j) is 2*Ne*(f[i] + f[j])*delta[i, j];
# with flat f this is 4*Ne*delta, so a coefficient of S corresponds to:
delta <- 1 + S / (4 * Ne)
```

The intermediate depth sets how tightly the two substitutions couple: the
rate of escaping an intermediate grows roughly like $e^{|2Ns|}$, while
non-synonymous scaling (below) pins the overall rate of excursions. A
deeper intermediate therefore gives the same number of toggles, each
resolved much faster.

## Model

Non-synonymous scaling makes one unit of branch length one expected
amino-acid-changing substitution per codon, so the compensatory events are
delivered at a rate the tree's branch lengths control directly.

```{r}
hky <- HasegawaKishinoYano(equilibrium = rep(0.25, 4))

co <- CoEvolution(
    population_size = Ne,
    mutation_rate = 1e-6,
    nucleotide_model = hky,
    fitness_1 = f,
    fitness_2 = f,
    delta = delta,
    scaling_type = "non-synonymous")
```

At equilibrium the two compatible configurations hold essentially all the
mass:

```{r}
eq <- as.numeric(co$equilibrium)
sum(sort(eq, decreasing = TRUE)[1:8])   # the 8 codon pairs spelling K,E / E,K
```

## Simulation

```{r}
s <- sample_sequence(model = co, length = 30)

sim <- simulate_over_phylogeny(
    phylogeny = mammals,
    model = co,
    sequence = s)
```

The interaction keeps the pair under strong purifying selection, so almost
every event is synonymous churn within the K and E codon families; the
compensatory dynamics live in the small non-synonymous remainder:

```{r}
nrow(sim$substitutions)
ns <- subset(sim$substitutions, !(first_synonymous & second_synonymous))
nrow(ns)
```

Substitutions into the two intermediates (excursion starts) are balanced by
non-synonymous substitutions back into the compatible configurations
(excursion ends), and no other amino-acid pair is ever reached:

```{r}
sort(table(ns$pair_amino_acid_to), decreasing = TRUE)
```

Each excursion starts with a substitution into K,K or E,E and resolves with
the next non-synonymous substitution at the same pair -- either completing
the toggle (landing on the opposite compatible configuration) or reverting.
A small helper classifies each excursion and records how long the
intermediate lasted:

```{r}
count_toggles <- function(substitutions, compatible = c("K,E", "E,K")) {
  toggles <- 0; reversions <- 0; waits <- c()
  for (s0 in unique(substitutions$site)) {
    for (nd in unique(substitutions$node)) {
      g <- substitutions[substitutions$site == s0 & substitutions$node == nd, ]
      if (nrow(g) < 2) next
      g <- g[order(g$time), ]
      for (i in seq_len(nrow(g) - 1)) {
        entered  <- !(g$pair_amino_acid_to[i] %in% compatible)
        resolved <- g$pair_amino_acid_to[i + 1] %in% compatible
        if (entered && resolved) {
          waits <- c(waits, g$time[i + 1] - g$time[i])
          if (g$pair_amino_acid_from[i] != g$pair_amino_acid_to[i + 1]) {
            toggles <- toggles + 1
          } else {
            reversions <- reversions + 1
          }
        }
      }
    }
  }
  list(toggles = toggles, reversions = reversions, waits = waits)
}

excursions <- count_toggles(ns)
excursions$toggles
excursions$reversions
median(excursions$waits)
```

Completed toggles and reversions occur in the expected 1:1 ratio, and the
median wait in the intermediate is well under one percent of a typical
branch length -- the compensating substitution follows essentially
immediately.

## Substitution history

Plotting every event would bury the signal in synonymous churn, so plot
only the non-synonymous events:

```{r}
sim_ns <- sim
sim_ns$substitutions <- ns
plot(sim_ns)
```

Each event draws one dot for the member of the pair that changed: a dot
above the branch is a substitution at the first site, a dot below is a
substitution at the second (green synonymous, red non-synonymous -- here
every plotted event is non-synonymous). The compensated toggles are the
tight vertical pairs of red dots, one above and one below the branch at
almost the same position: one site substitutes and the other follows
within a fraction of a percent of the branch length.

The entry--exit structure is visible directly in the filtered substitution
table:

```{r, results='hide'}
head(ns)
```
```{r, echo=F, results='asis'}
kable(head(ns), row.names = FALSE)
```

## Alignment

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```

Show alignment in a scrollable viewer:

```{r}
plot(sim$alignment)
```
