---
output:
  html_document:
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
set_palantir_seed(20260820)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Co-Evolution

Co-Evolution is an extension of a Mutation-Selection codon model, which considers pairs of codons.

The model adds an amino-acid correlation matrix $\delta$, which expresses how "linked" two amino acid states are within a pair. 

For example, $\delta_{A, K} > 1$ means that $AK$ is a favoured pair, while $\delta_{W, Y} < 1$ means $WY$ is disfavoured. $\delta = 1$ signifies lack of association.

The selection coefficient is calculated for a pair of sites, assuming additive fitness.

For a substitution from codon pair $ij$ to codon pair $kj$ (a change at the
first site; changes at the second site are symmetric):

$$
s_{ij, kj} = (\psi_k + \psi_j)\delta_{kj} - (\psi_i + \psi_j)\delta_{ij} \\
Q_{ij, kj} = 2N_e \, \mu_{ik} \, \frac{1-e ^ {-s_{ij, kj}}}{1-e^{-2 N_e s_{ij, kj}}}
$$

Only single-nucleotide codon changes at a single site of the pair have
non-zero instantaneous rate; simultaneous changes at both sites have rate
zero (as of version 1.2.0), so compensated pairs arise as two rapid
sequential substitutions.

* $N_e$ is the effective population size
* $\mu_{ij}$ is the mutation rate between a pair of nucleotides. This rate is taken from a nucleotide substitution model, such as Hasegawa Kishino Yano [[1](https://www.ncbi.nlm.nih.gov/pubmed/3934395)].
* $\psi_i$ is the absolute fitness of amino acid for codon $i$. Values of $\psi$ for the first and second pair can be specified from different fitness vectors.
* $s_{ij, kl}$ is selection coefficient for a pair substitution. If either substitution in a pair is synonymous, the second term of the product will be equal to $1$.

# Example: compensatory evolution at a salt bridge

We construct a pair landscape in which only two amino-acid configurations
have high fitness: lysine--glutamate and glutamate--lysine, the classic
charge-swap of an electrostatic interaction. The single-mismatch
configurations (K,K and E,E) are mildly deleterious, so the pair can pass
through them, and every other configuration is strongly deleterious. With
this landscape the pair should toggle between K,E and E,K through brief
intermediates: a substitution at one site followed rapidly by the
compensating substitution at the other.

## Setup

```{r}
# Load phylogeny
mammals <- Phylogeny("../inst/extdata/mammals.newick")

# Flat site fitness: all structure comes from the pair coupling
f <- rep(1, 20)

# Amino acid order is:
# A, C, D, E, F, G, H, I, K, L, M, N, P, Q, R, S, T, V, W, Y
K <- 9; E <- 4

# Three-tier coupling. With fitness_1 + fitness_2 = 2 and N = 1000, a pair
# with delta = 1 - g/4000 sits a scaled gap of 2*N*2*(g/4000) = g below the
# compatible pairs:
delta <- matrix(1 - 20/4000, 20, 20)   # everything else:      gap 20
delta[K, K] <- 1 - 4/4000              # mismatch K,K:         gap  4
delta[E, E] <- 1 - 4/4000              # mismatch E,E:         gap  4
delta[K, E] <- 1                       # compatible:           gap  0
delta[E, K] <- 1
```

## Model

Non-synonymous scaling makes one unit of branch length one expected
amino-acid-changing substitution per codon, so the compensatory events are
delivered at a rate the tree's branch lengths control directly.

```{r}
hky <- HasegawaKishinoYano(equilibrium = rep(0.25, 4))

co <- CoEvolution(
    population_size = 1000,
    mutation_rate = 1e-6,
    nucleotide_model = hky,
    fitness_1 = f,
    fitness_2 = f,
    delta = delta,
    scaling_type = "non-synonymous")
```

At equilibrium the two compatible configurations hold essentially all the
mass:

```{r}
eq <- as.numeric(co$equilibrium)
sum(sort(eq, decreasing = TRUE)[1:8])   # the 8 codon pairs spelling K,E / E,K
```

## Simulation

```{r}
s <- sample_sequence(model = co, length = 30)

sim <- simulate_over_phylogeny(
    phylogeny = mammals,
    model = co,
    sequence = s)
```

The pair states visited confirm the design -- the two compatible
configurations dominate, with the designed intermediates and nothing else:

```{r}
sort(table(sim$substitutions$pair_amino_acid_to), decreasing = TRUE)
```

Count the completed toggles between the two compatible configurations:

```{r}
toggles <- 0
for (s0 in unique(sim$substitutions$site)) {
  for (nd in unique(sim$substitutions$node)) {
    g <- subset(sim$substitutions, site == s0 & node == nd)
    if (!nrow(g)) next
    states <- g$pair_amino_acid_to[order(g$time)]
    states <- rle(states[states %in% c("K,E", "E,K")])$values
    if (length(states) > 1) toggles <- toggles + sum(states[-1] != states[-length(states)])
  }
}
toggles
```

## Substitution history

Plot substitution history:

```{r}
plot(sim, sites = 0:7)
```

Each substitution event is drawn as two stacked dots: the top dot colours
the first site of the pair, the bottom dot the second (green synonymous,
red non-synonymous). The compensated toggles appear as red marks in close
succession on a branch -- a top-red event at one site followed almost
immediately by a bottom-red event at the other, as the pair passes through
its brief intermediate.

Check substitution history table:

```{r, results='hide'}
head(sim$substitutions)
```
```{r, echo=F, results='asis'}
kable(head(sim$substitutions))
```

## Alignment

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```

Show alignment in a scrollable viewer:

```{r}
plot(sim$alignment)
```
