---
title: "Goldman--Yang (GY94) codon models"
output:
  html_document:
    self_contained: no
---

```{r, include=FALSE}
library(PalantiR)
library(knitr)
set_palantir_seed(20260825)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Model

The Goldman--Yang 1994 (GY94) model describes codon substitutions using a
transition/transversion rate ratio, `kappa`, and a non-synonymous/synonymous
rate ratio, `omega`. For distinct sense codons (i) and (j), PalantiR first
constructs the unscaled generator

$$
\widetilde q_{ij} =
\begin{cases}
0, & \text{if }i\text{ and }j\text{ differ at more than one nucleotide},\\
\pi_j, & \text{synonymous transversion},\\
\kappa\pi_j, & \text{synonymous transition},\\
\omega\pi_j, & \text{non-synonymous transversion},\\
\omega\kappa\pi_j, & \text{non-synonymous transition}.
\end{cases}
$$

Here (pi_j) is the stationary frequency of target codon (j). Diagonal
entries are the negative row sums. Both `omega` and `kappa` must be finite and
strictly positive. The familiar interpretation is `omega < 1` for purifying
selection, `omega = 1` for neutral amino-acid evolution, and `omega > 1` for an
excess of non-synonymous change, although the constructor does not impose those
biological categories.

```{r}
gy <- GoldmanYang94(
    omega = 0.2,
    kappa = 2,
    frequencies = c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    frequency_model = "F1x4",
    scaling_type = "standard")

gy$omega
gy$scaling_type
```

`GY94()` is a short alias for `GoldmanYang94()`.

# Stationary codon frequencies

All frequency adapters operate on the sense codons of the active genetic code.
They remove stop codons and normalize the resulting sense-codon probabilities
to sum to one.

* `FEqual()` assigns equal probability to every sense codon.
* `F1x4()` accepts four nucleotide frequencies shared by all three codon
  positions.
* `F3x4()` accepts a 3 by 4 matrix (positions by nucleotides), its 4 by 3
  transpose, or a length-12 vector in three position blocks.
* `F61()` accepts one positive value per sense codon. The name is conventional:
  the actual length follows the active genetic code. Named input is safest and
  must contain every sense codon exactly once.

The nucleotide order for unnamed input is `T, C, A, G`. Named input may be in
any order and is rearranged internally.

```{r}
pi_equal <- FEqual()
pi_1x4 <- F1x4(c(T = 0.30, C = 0.20, A = 0.25, G = 0.25))

position_frequencies <- rbind(
    c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    c(T = 0.20, C = 0.30, A = 0.30, G = 0.20),
    c(T = 0.25, C = 0.25, A = 0.20, G = 0.30))
pi_3x4 <- F3x4(position_frequencies)

f61_input <- setNames(rep(1, length(GeneticCode())), names(GeneticCode()))
pi_61 <- F61(f61_input)

sum(pi_3x4)
head(pi_3x4)
```

`codon_frequencies()` performs the same adaptation and can infer F1x4, F3x4,
or F61 from input length. An explicit `frequency_model` is recommended in
reproducible scripts. A `CodonFrequencies` object records the active genetic
code; PalantiR rejects it if the genetic code is subsequently changed, because
the codon state ordering may no longer agree.

# Branch-length scaling

The unscaled matrix is divided by

$$
S_t = \sum_i \pi_i \sum_{j\in t}\widetilde q_{ij},
$$

where (t) is the selected event class:

* `"standard"` or `"substitution"`: all codon substitutions;
* `"non-synonymous"`: amino-acid-changing substitutions only;
* `"synonymous"`: synonymous substitutions only;
* `"none"`: no normalization.

`"standard"` is an input alias for `"substitution"`; the model stores the
canonical value `"substitution"`. With a scaled model, one branch-length unit
means one expected event of the selected class per codon at that model's
stationary distribution.

In a discrete site mixture, **each class is scaled independently**. A branch of
length 0.1 therefore represents 0.1 expected events of the selected class per
site within every site class at its own stationary distribution. PalantiR does
not compute one mixture-wide normalization across classes. All branch modes
within one time-heterogeneous class must use the same scaling type.

# Branch heterogeneity

`GY94BranchModel()` uses the same mode-tree convention as other PalantiR
time-heterogeneous models. Mode labels are zero-based indices into `models`, and
`start_mode` specifies the process at the root. Any GY94 parameter may change
between modes, including `omega`, `kappa`, and the stationary codon frequencies.

```{r}
mammals <- Phylogeny("../inst/extdata/mammals.newick")
mammal_modes <- Phylogeny("../inst/extdata/mammals_switch.newick", type = "mode")

background <- GY94(omega = 0.2, kappa = 2, frequencies = pi_1x4)
foreground <- GY94(omega = 3.0, kappa = 2, frequencies = pi_3x4)

episodic <- GY94BranchModel(
    models = list(background, foreground),
    mode_phylogeny = mammal_modes,
    start_mode = 0)
```

When stationary frequencies change, descendants enter the new process with the
state distribution inherited from the parent; they are not redrawn from the
new equilibrium. PalantiR simulates the resulting transient and uses the
selected temporal rescaling method to preserve the chosen branch-length
currency. The exact time change is the default; the segment approximation
remains available through `rescale_method = "segments"`.

# Arbitrary discrete site classes

`GY94SiteModel()` accepts any collection of homogeneous `GY94` models and
time-heterogeneous `GY94BranchModel` specifications. Supply either `n_sites`,
one count per class, or `site_classes`, one assignment per output position.
Character assignments refer to class labels; numeric assignments are one-based
indices into `models`. Output `site` and `class_index` fields are zero-based.

```{r}
site_model <- GY94SiteModel(
    models = list(
        constrained = background,
        neutral = GY94(omega = 1, kappa = 2, frequencies = pi_1x4),
        episodic = episodic),
    n_sites = c(40, 30, 30),
    labels = c("constrained", "neutral", "episodic"))
```

For an explicit interleaved assignment, use for example
`site_classes = c("constrained", "episodic", "neutral", ...)`; the simulator
restores that requested order after efficiently simulating sites in class
blocks.

# Zhang--Nielsen--Yang branch-site model

`ZNYBranchSiteModel()` builds the four standard branch-site classes as a
special case of the general site-class representation:

| Site class | Background omega | Foreground omega |
|:----------:|:----------------:|:----------------:|
| 0          | `omega0`         | `omega0`         |
| 1          | 1                | 1                |
| 2a         | `omega0`         | `omega2`         |
| 2b         | 1                | `omega2`         |

The mode tree must label background branches `0` and foreground branches `1`.
`site_counts` gives counts in the order 0, 1, 2a, 2b. Foreground frequencies
default to the background frequencies, but may be supplied separately when the
stationary codon distribution also changes on foreground branches.

```{r}
branch_site <- ZNYBranchSiteModel(
    mode_phylogeny = mammal_modes,
    site_counts = c(4, 4, 4, 4),
    omega0 = 0.2,
    omega2 = 3,
    kappa = 2,
    frequencies = c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    frequency_model = "F1x4",
    scaling_type = "standard")
```

`ZNYBranchSite()` is a short alias.

# Simulation and output metadata

The simulator samples a root state for each site from the equilibrium of that
class's root model unless a codon `Sequence` is supplied explicitly.

```{r, eval=FALSE}
sim <- simulate_gy94_site_model(mammals, branch_site)

head(sim$site_classes)
head(sim$substitutions)
plot(sim$alignment)
plot(sim, sites = 0:9)
```

For the class counts above, `sim$site_classes` begins:

```
  site site_class class_index
1    0          0           0
2    1          0           0
3    2          0           0
4    3          0           0
5    4          1           1
6    5          1           1
```

The complete assignment table is stored in `sim$site_classes`, including sites
with no substitutions. The same table is attached to `sim$alignment`; the
interactive alignment header displays the class label. Every event in
`sim$substitutions` carries `site_class` and `class_index`, which also appear in
substitution-history tooltips. `sim$site_model` retains the general class
specification and `sim$site_class_models` maps labels to their homogeneous or
branch-heterogeneous model definitions.

# References

1. Goldman, N. and Yang, Z. (1994). A codon-based model of nucleotide
   substitution for protein-coding DNA sequences. *Molecular Biology and
   Evolution* **11**, 725--736.
2. Zhang, J., Nielsen, R. and Yang, Z. (2005). Evaluation of an improved
   branch-site likelihood method for detecting positive selection at the
   molecular level. *Molecular Biology and Evolution* **22**, 2472--2479.
