---
title: "Goldman--Yang (GY94) codon models"
output:
  html_document:
    self_contained: no
---

```{r, include=FALSE}
library(PalantiR)
library(knitr)
set_palantir_seed(20260825)
knitr::opts_chunk$set(fig.height = 7, fig.width = 9)
```

# Model

The Goldman--Yang 1994 (GY94) model describes codon substitutions using a
transition/transversion rate ratio, `kappa`, and a non-synonymous/synonymous
rate ratio, `omega`. For distinct sense codons (i) and (j), PalantiR first
constructs the unscaled generator

$$
\widetilde q_{ij} =
\begin{cases}
0, & \text{if }i\text{ and }j\text{ differ at more than one nucleotide},\\
\pi_j, & \text{synonymous transversion},\\
\kappa\pi_j, & \text{synonymous transition},\\
\omega\pi_j, & \text{non-synonymous transversion},\\
\omega\kappa\pi_j, & \text{non-synonymous transition}.
\end{cases}
$$

Here (pi_j) is the stationary frequency of target codon (j). Diagonal
entries are the negative row sums. Both `omega` and `kappa` must be finite and
strictly positive. The familiar interpretation is `omega < 1` for purifying
selection, `omega = 1` for neutral amino-acid evolution, and `omega > 1` for an
excess of non-synonymous change, although the constructor does not impose those
biological categories.

```{r}
use_genetic_code("Standard nuclear")

gy <- GoldmanYang94(
    omega = 0.2,
    kappa = 2,
    frequencies = c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    frequency_model = "F1x4",
    scaling_type = "standard")

gy$omega
gy$scaling_type
```

`GY94()` is a short alias for `GoldmanYang94()`.

# Stationary codon frequencies

All frequency adapters operate on the sense codons of the active genetic code.
They remove stop codons and normalize the resulting sense-codon probabilities
to sum to one.

* `FEqual()` assigns equal probability to every sense codon.
* `F1x4()` accepts four nucleotide frequencies shared by all three codon
  positions.
* `F3x4()` accepts a 3 by 4 matrix (positions by nucleotides), its 4 by 3
  transpose, or a length-12 vector in three position blocks.
* `F61()` accepts one positive value per sense codon. The name is conventional:
  the actual length follows the active genetic code. Named input is safest and
  must contain every sense codon exactly once.

The nucleotide order for unnamed input is `T, C, A, G`. Named input may be in
any order and is rearranged internally.

```{r}
use_genetic_code("Standard nuclear")

pi_equal <- FEqual()
pi_1x4 <- F1x4(c(T = 0.30, C = 0.20, A = 0.25, G = 0.25))

position_frequencies <- rbind(
    c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    c(T = 0.20, C = 0.30, A = 0.30, G = 0.20),
    c(T = 0.25, C = 0.25, A = 0.20, G = 0.30))
pi_3x4 <- F3x4(position_frequencies)

f61_input <- setNames(rep(1, length(GeneticCode())), names(GeneticCode()))
pi_61 <- F61(f61_input)

sum(pi_3x4)
head(pi_3x4)
```

`codon_frequencies()` performs the same adaptation and can infer F1x4, F3x4,
or F61 from input length. An explicit `frequency_model` is recommended in
reproducible scripts. A `CodonFrequencies` object records the active genetic
code; PalantiR rejects it if the genetic code is subsequently changed, because
the codon state ordering may no longer agree.

# Branch-length scaling

The dS choice is handled through a neutral reference. For `"synonymous"`,
`"dS"`, or `"ds"`, PalantiR constructs the `omega = 1` generator
\(\widetilde Q^{(1)}\) and uses the denominator

$$
S_{dS}=-\frac{1}{3}\sum_i\pi_i\widetilde q^{(1)}_{ii}.
$$

The selected generator is divided by \(S_{dS}\), so the neutral total rate is
three substitutions per codon and branch length is dS per synonymous
opportunity. The denominator is independent of the selected `omega`, and the
model stores the canonical scaling type `"dS"`.

The other scaled choices are event-count currencies. Let \(\mathcal A\) be the
set of ordered codon transitions that determines the count. For one GY94
process, the corresponding stationary event rate is

$$
S_{\mathcal A} = \sum_i \pi_i
\sum_{j:(i,j)\in\mathcal A}\widetilde q_{ij}.
$$

The event-count choices for \(\mathcal A\) are:

* `"standard"` or `"substitution"`: all codon substitutions;
* `"non-synonymous"`: amino-acid-changing substitutions only;
* `"synonymous-per-codon"`: synonymous substitutions only;
* `"none"`: no normalization.

`"standard"` is an input alias for `"substitution"`; the model stores the
canonical value `"substitution"`. A standalone event-count model uses
\(Q=\widetilde Q/S_{\mathcal A}\), so one branch-length unit represents one
expected event from \(\mathcal A\) per codon at stationarity.

For a discrete mixture with class proportions \(p_k\), PalantiR follows the
mixture-wide convention used for codon site models in PAML:

$$
\overline S_{\mathcal A}=\sum_k p_k S_{\mathcal A,k},
\qquad
Q_k=\widetilde Q_k/\overline S_{\mathcal A}.
$$

Every component therefore shares one denominator, while its stationary rate
\(S_{\mathcal A,k}/\overline S_{\mathcal A}\) is retained. The
mixture-weighted mean of those rates is one; individual component rates need
not be one. Under synonymous-per-codon or non-synonymous scaling, independent
normalization would erase omega-dependent rate differences among classes.
The dS mixture uses the same shared-denominator construction with
class-specific neutral denominators. In the usual GY94 site and branch-site
models, `kappa` and codon frequencies are shared within a branch type, so every
class has the same dS denominator exactly.

For a branch-site model, PalantiR computes the weighted denominator separately
for each branch type, because the collection of active class-specific
generators differs between background and foreground branches:

$$
\overline S_{\mathcal A,b}=\sum_k p_k S_{\mathcal A,k,b}.
$$

All site classes active on branch type \(b\) share
\(\overline S_{\mathcal A,b}\). Thus, under standard scaling, branch length is
the expected number of codon substitutions per site averaged over the whole
mixture. Synonymous-per-codon and non-synonymous scaling give the analogous mean for the
specified transition set. This is the convention implemented for site and
branch-site models in the [PAML `codeml`
source](https://github.com/abacus-gene/paml/blob/master/src/codeml.c#L2580-L2629),
which notes that scaling applies to all class-specific matrices for the whole
branch. All classes and branch modes within one mixture must use the same
scaling type.

# Branch heterogeneity

`GY94BranchModel()` uses the same mode-tree convention as other PalantiR
time-heterogeneous models. Mode labels are zero-based indices into `models`, and
`start_mode` specifies the process at the root. Any GY94 parameter may change
between modes, including `omega`, `kappa`, and the stationary codon frequencies.

```{r}
mammals <- Phylogeny("../inst/extdata/mammals.newick")
mammal_modes <- Phylogeny("../inst/extdata/mammals_switch.newick", type = "mode")

background <- GY94(omega = 0.2, kappa = 2, frequencies = pi_1x4,
                   scaling_type = "standard")
foreground <- GY94(omega = 3.0, kappa = 2, frequencies = pi_3x4,
                   scaling_type = "standard")

episodic <- GY94BranchModel(
    models = list(background, foreground),
    mode_phylogeny = mammal_modes,
    start_mode = 0)
```

When stationary frequencies change, descendants enter the new process with the
state distribution inherited from the parent; they are not redrawn from the
new equilibrium. PalantiR uses the requested temporal rescaling method to
preserve each component's branch-length budget relative to the common mixture
denominator. For event-count gauges this is the requested class-event outflux.
For dS it is one third of the per-state total outflux from the `omega = 1`
reference, integrated under the evolving codon distribution. Thus dS remains
neutral nucleotide time without being converted into a fixed realized
synonymous-event count. The exact time change is the default; the segment
approximation remains available through `rescale_method = "segments"`.

# Arbitrary discrete site classes

`GY94SiteModel()` accepts any collection of homogeneous `GY94` models and
time-heterogeneous `GY94BranchModel` specifications. Supply either `n_sites`,
one count per class, or `site_classes`, one assignment per output position.
Character assignments refer to class labels; numeric assignments are one-based
indices into `models`. Branch-heterogeneous classes in one mixture must use the
same mode phylogeny and number of branch models, because their generators are
normalized together on each branch type. Output `site` and `class_index` fields
are zero-based.

```{r}
site_model <- GY94SiteModel(
    models = list(
        constrained = background,
        neutral = GY94(omega = 1, kappa = 2, frequencies = pi_1x4,
                       scaling_type = "standard"),
        episodic = episodic),
    n_sites = c(40, 30, 30),
    labels = c("constrained", "neutral", "episodic"))
```

For an explicit interleaved assignment, use for example
`site_classes = c("constrained", "episodic", "neutral", ...)`; the simulator
restores that requested order after efficiently simulating sites in class
blocks.

# Zhang--Nielsen--Yang branch-site model

`ZNYBranchSiteModel()` builds the four standard branch-site classes as a
special case of the general site-class representation:

| Site class | Background omega | Foreground omega |
|:----------:|:----------------:|:----------------:|
| 0          | `omega0`         | `omega0`         |
| 1          | 1                | 1                |
| 2a         | `omega0`         | `omega2`         |
| 2b         | 1                | `omega2`         |

The mode tree must label background branches `0` and foreground branches `1`.
`site_counts` gives counts in the order 0, 1, 2a, 2b. Foreground frequencies
default to the background frequencies, but may be supplied separately when the
stationary codon distribution also changes on foreground branches.

```{r}
branch_site <- ZNYBranchSiteModel(
    mode_phylogeny = mammal_modes,
    site_counts = c(4, 4, 4, 4),
    omega0 = 0.2,
    omega2 = 3,
    kappa = 2,
    frequencies = c(T = 0.30, C = 0.20, A = 0.25, G = 0.25),
    frequency_model = "F1x4",
    scaling_type = "standard")
```

`ZNYBranchSite()` is a short alias.

# Simulation and output metadata

The simulator samples a root state for each site from the equilibrium of that
class's root model unless a codon `Sequence` is supplied explicitly.

```{r, eval=FALSE}
sim <- simulate_gy94_site_model(mammals, branch_site)

head(sim$site_classes)
head(sim$substitutions)
plot(sim$alignment)
plot(sim, sites = 0:9)
```

For the class counts above, `sim$site_classes` begins:

```
  site site_class class_index
1    0          0           0
2    1          0           0
3    2          0           0
4    3          0           0
5    4          1           1
6    5          1           1
```

The complete assignment table is stored in `sim$site_classes`, including sites
with no substitutions. The same table is attached to `sim$alignment`; the
interactive alignment header displays the class label. Every event in
`sim$substitutions` carries `site_class` and `class_index`, which also appear in
substitution-history tooltips. `sim$site_model` retains the general class
specification and `sim$site_class_models` maps labels to their homogeneous or
branch-heterogeneous model definitions. The generating `GY94SiteModel` also
records `class_weights`, the `mixture_scaling` denominator for each branch
mode, and `stationary_class_rates` relative to that denominator.

For a complete executable example, the
[live R/PalantiR history-validation notebook](../validation/gy94_deep_tree_history_dnds_R.ipynb)
constructs a 50-taxon branch-heterogeneous GY94 model, simulates 5,000 codon
sites, explores the native interactive viewers, and recovers branch-specific
$d_N/d_S$ directly from all recorded substitution histories.

# References

1. Goldman, N. and Yang, Z. (1994). A codon-based model of nucleotide
   substitution for protein-coding DNA sequences. *Molecular Biology and
   Evolution* **11**, 725--736.
2. Yang, Z., Nielsen, R., Goldman, N. and Pedersen, A.-M. K. (2000).
   Codon-substitution models for heterogeneous selection pressure at amino
   acid sites. *Genetics* **155**, 431--449.
3. Yang, Z. and Nielsen, R. (2002). Codon-substitution models for detecting
   molecular adaptation at individual sites along specific lineages.
   *Molecular Biology and Evolution* **19**, 908--917.
4. Zhang, J., Nielsen, R. and Yang, Z. (2005). Evaluation of an improved
   branch-site likelihood method for detecting positive selection at the
   molecular level. *Molecular Biology and Evolution* **22**, 2472--2479.
