---
output: 
  html_document: 
    self_contained: no
---

```{r, include=F}
library(PalantiR)
library(knitr)
```

# Mutation-Selection

Mutation Selection (MS) model ([1](https://www.ncbi.nlm.nih.gov/pubmed/18178545), [2](http://link.springer.com/article/10.1007/BF02102078)) is the core model implemented in `PalantiR`.
Mutation Selection is a codon substitution model that describes the interplay between mutation rate, selection strength, and population size.

For a substitution from codon $i$ to codon $j$:
$$
s_{ij} = \psi_j - \psi_i \\
Q_{ij} = 2N_e \mu_{ij} \frac{1-e^{s_{ij}}}{1-e^{-2N_es_{ij}}}
$$
Only codons differing in one nucleotide positions are considered. Rates of transitions for multiple codon differences are $0$.

* $N_e$ is the effective population size
* $\mu_{ij}$ is the mutation rate between a pair of nucleotides. This rate is taken from a nucleotide substitution model, such as Hasegawa Kishino Yano [[1](https://www.ncbi.nlm.nih.gov/pubmed/3934395)].
* $\psi_i$ is the absolute fitness of amino acid for codon $i$.
* $s_{ij}$ is selection coefficient of a substitution. For synonymous substitutions, the second term of the product will be equal to $1$.

The substitution rates are also scaled to assign particular meaning to branch length of a phylogeny. For more detail, see [[Rate Scaling]].

# Example

## Setup

Load model parameters:

```{r}
# Load phylogeny
mammals <- Phylogeny("../data/mammals.newick")

# Load nucleotide equilibrium distribution
# Nucleotide order is T,C,A,G
nuc_pi <- c(.3, .2, .25, .25)

# Load amino acid fitness vector
# Amino acid order is:
# A, C, D, E, F, G, H, I, K, L, M, N, P, Q, R, S, T, V, W, Y
aa_psi <- as.matrix(read.csv("../data/amino_acid_fitness_N_1000.csv"))
```

Create model:

```{r}
# Create nucleotide substitution model
hky <- HasegawaKishinoYano(nuc_pi)

# Create codon substitution model
ms <- MutationSelection(
    population_size = 1000,
    mutation_rate = 1e-8,
    nucleotide_model = hky,
    fitness = aa_psi[1,])
```

Sample sequence from model equilibrium distribution to start at root:

```{r}
s <- sample_sequence(model = ms, length = 100)
```

Simulate:

```{r}
sim <- simulate_over_phylogeny(
    phylogeny = mammals,
    model = ms,
    sequence = s)
```

## Substitution history

Plot substitution history:

```{r}
plot(sim, sites = 0:10)
```

Synonymous substiutions are shown in green, non-synonymous in red.

Check substitution history table:

```{r, results='hide'}
head(sim$substitutions)
```
```{r, echo=F, results='asis'}
kable(head(sim$substitutions))
```

## Alignment

Show alignment in a viewer:

```{r}
plot(sim$alignment)
```

To export `FASTA` sequence:

```{r, eval=F}
as.fasta(sim$alignment, file = "palantir_sim.fa")
```


# References

1. Golding, B. and Felsenstein, J. (1990). _A maximum likelihood approach to the detection of selection from a phylogeny_. Journal of Molecular Evolution, 31(6):511–523.
2. Yang, Z. and Nielsen, R. (2008). _Mutation-Selection Models of Codon Substitution and Their Use to Estimate
Selective Strengths on Codon Usage_. Molecular Biology and Evolution. 25(3):568-579.
3. Hasegawa, M., Kishino, H. and Yano, T. (1985). _Dating of the human-ape splitting by a molecular clock of mitochondrial DNA_. Journal of Molecular Evolution 22:160-174.
